Tissue Enrichment Analysis for C. elegans Genomics

### 2016-04-25 tests/hgf_benchmarking/doc/tea_main_text.tex
uploading paper and updating ipy notebook

\title{Tissue Enrichment Analysis (TEA)}

@@

\begin{abstract} % abstract
\parttitle{Background} %if any
Over the last ten years, there has been explosive development in tools
for measuring gene expression. These tools can identify thousands of altered 
genes between conditions, but understanding these datasets and forming hypotheses based on them remains challenging. One way to analyze these datasets is to associate ontologies (hierarchical, descriptive vocabularies
with controlled relations between terms) with genes and to look for enrichment of specific terms. Although gene ontology (GO) is available for   \emph{Caenorhabditis~elegans}, it does not include anatomical information.
\parttitle{Results} %if any 
We have developed a tool for identifying  enrichment of \emph{C.~elegans} tissues among gene sets and generated a website GUI where users can access this tool. Since a common drawback to ontology enrichment analyses is result verbosity, we developed a very simple trimming algorithm to reduce the ontology size by an order of magnitude. We adjusted these filters and validated our tool using a set of 30  gold standards from Expression Cluster data in WormBase and we show our tool can even discriminate between embryonic and larval tissues and can even identify tissues down to the single-cell level. We used our tool to identify multiple neuronal tissues that are down-regulated due to pathogen infection in \emph{C.~elegans}.
\parttitle{Conclusions} %if any
Our Tissue Enrichment Analysis (TEA) can be found at \url{www.wormbase.org/tea}, and can be downloaded using Python's standard pip installer. It tests a slimmed-down \emph{C.~elegans} tissue ontology for enrichment of specific terms and provides users with a  text and graphic representation of the results. 
\end{abstract}

@@

Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene. %There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (pers. comm., H. Yu and P. Thomas). To reduce verbosity, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information. To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms. Other pruning tools do exist (see for example~\cite{Kim2007}), but the pruning is query-dependent. We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.

@@

\subsubsection*{Terminal branch terms and parent terms can be
safely removed in an algorithmic fashion }
Another problem arises from the ontology being scarcely populated. Many nodes have 0-10 annotations, which we consider too few to accurately test. To solve this issue, we implemented a straightforward trimming strategy. For a given terminal node, we test whether the node has more than a threshold number of annotations. If it does not, the node is removed. The next higher node in the branch is tested and removed recursively until a node which satisfies the condition is found. At that point, no more nodes can be removed from that branch. This is guaranteed by the structure of the ontology: parent nodes inherit all of the annotations of all of their descendants, so the number of annotated terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}). In this way, we ensure that our term dictionary includes only those tissues that are considered sufficiently well annotated for statistical purposes.

@@

%Other tools
Our tool relies on an annotation dictionary that is continuously updated primarily with data from single gene qualitative analyses, does not require retraining and does not require ranked genes. To our knowledge, this is the first tool that tests tissue enrichment  in \emph{C.~elegans} via the hypergeometric method, but similar projects exist for humans and zebrafish~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology. Chikina \emph{et al}~\cite{Chikina2009} have previously reported a tissue enrichment model for \emph{C.~elegans }based on a Support Vector Machine classifier that has been trained on microarray studies. SVMs are powerful tools, but they require continuous retraining as more tissue expression data becomes available. Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for certain studies. Finally, this tissue enrichment tool provides users with enrichment results for only 6 large tissues. In contrast, our tool routinely tests a much larger number of terms, and we have shown it can even accurately identify enrichment of embryonic precursor lineages for select data sets.

@@

\section*{Availability of data and materials}
Our web implementation is available at \url{https://www.wormbase.org/tea}. Our software can also be downloaded using Python's pip installer via the command

@@

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
% \nocite{label}

@@

% statistical testing and validation Figures
%q plot
\begin{figure}[h!]
	\includegraphics[width=\textwidth]{avgQKDE_method=any.pdf}
  \captionsetup{width= 0.95\textwidth}
  \caption{
  \csentence{Kernel density estimates for 30 gold standard datasets.}
      We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values. The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.
	  }
	  \label{fig:qvals}
\end{figure}

@@

\begin{table}[h!]
	\caption{
	Comparison of results for a neuronal-enriched gene set from Watson~\cite{Watson2008a}. We ran the same genelist on a dictionary with a minimum annotation cutoff of 50, similarity threshold of 0.95 and similarity method `any' versus another with a minimum annotation cutoff of 33, similarity threshold of 0.95 and similarity method `any'. In the table, columns are labeled with their significance value (Q-value) or enrichment fold change followed by a hyphen and a number which indicates which the cutoff for the dictionary that was used for testing. Not all tissues are present in either dictionary. Hyphens denote not-applicable values, which occurs when a particular tissue is not present in both dictionaries. Full table is available on github.
	}

@@

\section*{Additional Files}
  \subsection*{Additional file 1 --- IPython Notebook}
    Tutorial for users interested in using our software within a python script

>>>>

\title{Tissue Enrichment Analysis for C. elegans Genomics}

@@

\begin{abstract} % abstract
\parttitle{Background} %if any
Over the last ten years, there has been explosive development in methods
for measuring gene expression. These methods can identify thousands of altered 
genes between conditions, but understanding these datasets and forming hypotheses based on them remains challenging. One way to analyze these datasets is to associate ontologies (hierarchical, descriptive vocabularies
with controlled relations between terms) with genes and to look for enrichment of specific terms. Although gene ontology (GO) is available for   \emph{Caenorhabditis~elegans}, it does not include anatomical information.
\parttitle{Results} %if any 
We have developed a tool for identifying  enrichment of \emph{C.~elegans} tissues among gene sets and generated a website GUI where users can access this tool. Since a common drawback to ontology enrichment analyses is result verbosity, we developed a very simple trimming algorithm to reduce the ontology size by an order of magnitude. We adjusted these filters and validated our tool using a set of 30  gold standards from Expression Cluster data in WormBase and we show our tool can even discriminate between embryonic and larval tissues and can even identify tissues down to the single-cell level. We used our tool to identify multiple neuronal tissues that are down-regulated due to pathogen infection in \emph{C.~elegans}.
\parttitle{Conclusions} %if any
Our Tissue Enrichment Analysis (TEA) can be found within WormBase, and can be downloaded using Python's standard pip installer. It tests a slimmed-down \emph{C.~elegans} tissue ontology for enrichment of specific terms and provides users with a  text and graphic representation of the results. 
\end{abstract}

@@

Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene. %There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (pers. comm., H. Yu and P. Thomas). To reduce verbosity, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information. To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms. Other pruning methods do exist (see for example~\cite{Kim2007}), but the pruning is query-dependent. We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.

@@

\subsubsection*{Terminal branch terms and parent terms can be
safely removed in an algorithmic fashion }
Another problem arises from the ontology being scarcely populated. Many nodes have 0-10 annotations, which we consider too few to accurately test. To solve this issue, we implemented a straightforward trimming strategy. For a given terminal node, we test whether the node has more than a threshold number of annotations. If it does not, the node is removed. The next higher node in the branch is tested and removed recursively until a node that satisfies the condition is found. At that point, no more nodes can be removed from that branch. This completion is guaranteed by the structure of the ontology: parent nodes inherit all of the annotations of all of their descendants, so the number of annotated terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}). In this way, we ensure that our term dictionary includes only those tissues that are considered sufficiently well annotated for statistical purposes.

@@

%Other tools
Our tool relies on an annotation dictionary that is continuously updated primarily with data from single gene qualitative analyses, does not require retraining and does not require ranked genes. To our knowledge, this is the first tool that tests tissue enrichment  in \emph{C.~elegans} via the hypergeometric method, but similar projects exist for humans and zebrafish~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology. Chikina \emph{et al}~\cite{Chikina2009} have previously reported a tissue enrichment model for \emph{C.~elegans }based on a Support Vector Machine classifier that has been trained on microarray studies. SVMs are powerful tools, but they require continuous retraining as more tissue expression data becomes available. Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for certain studies. Furthermore, this tissue enrichment tool provides users with enrichment results for only 6 large tissues. In contrast, our tool routinely tests a much larger number of terms, and we have shown it can even accurately identify enrichment of embryonic precursor lineages for select data sets.

@@

\section*{Availability of data and materials}
Our web implementation is available at \url{https://www.wormbase.org/enrichment/tea}. Our software can also be downloaded using Python's pip installer via the command

@@

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{tea_citations}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
% \nocite{label}

@@

% statistical testing and validation Figures
%q plot
\begin{figure}[h!]
	\includegraphics[width=\textwidth]{avgQKDE_method=any.pdf}
  \captionsetup{width= 0.95\textwidth}
  \caption{
  \csentence{Kernel density estimates (KDE) for 30 gold standard datasets.}
      We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values. The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.
	  }
	  \label{fig:qvals}
\end{figure}

@@

\begin{table}[h!]
	\caption{
	Comparison of results for a neuronal-enriched gene set from Watson~\cite{Watson2008a}. We ran the same genelist on a dictionary with a minimum annotation cutoff of 50, similarity threshold of 0.95 and similarity method `any' versus another with a minimum annotation cutoff of 33, similarity threshold of 0.95 and similarity method `any'. In the table, columns are labeled with their significance value (Q-value) or enrichment fold change followed by a hyphen and a number which indicates which the cutoff for the dictionary that was used for testing. Not all tissues are present in either dictionary. Hyphens denote not-applicable values, which occurs when a particular tissue is not present in both dictionaries. Full table is available in Additional File 2.
	}

@@

\section*{Additional Files}
  \subsection*{Additional file 1 --- IPython Notebook}
    Tutorial for users interested in using our software within a python script
  \subsection*{Additional file 2 --- Folder Structure for SI files 3 and 4}
    A file detailing the folder structure of the zipped folders 3 and 4. 
  \subsection*{Additional file 3 --- Golden Gene Sets}
    A list of all the genes used for our benchmarking process. 
  \subsection*{Additional file 4 --- Results}
    A folder containing a complete version of the results we generated for this paper.
